---
title: "Figure making in ggplot"
output: html_document
date: "13 February 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What we're doing today

In the spirit of Emily Riederer's [ugliest ggplot ever](https://gist.github.com/emilyriederer/2bf4f67d7e198f8359b61706c82e42ee), we'll play around with `ggplot` code in order to learn how it works. The goal: make the _ugliest_ plot possible.  

# Set up

We'll load in our packages below:
```{r message = FALSE, warning = FALSE}
library(tidyverse) # general tidying and visualization: ggplot is loaded by default if you use tidyverse
library(lterdatasampler) # data we're using comes from this package
library(lubridate) # working with dates
```

Note: if you don't have either the `lubridate` or `lterdatasampler` package, you'll have to install them before loading them in. `lubridate` is on CRAN, so you can do `install.packages("lubridate")`, but `lterdatasampler` has to be installed from the GitHub repo using the code below (copy, paste, and run in the console):
```{r eval = FALSE}
remotes::install_github("lter/lterdatasampler")
```

Today, we'll use the [Plum Island fiddler crab data](https://lter.github.io/lterdatasampler/articles/pie_crab_vignette.html) from `lterdatasampler` to visualize relationships between latitude and crab size. Read the linked vignette to learn about Bergmann's rule!  

The Plum Island LTER has a data set of crab sizes (column `size`) from Summer 2016 at 13 different marshes spanning 12 degrees latitude. A sample is below, but try `View(pie_crab)` in the console to see the whole data frame. 

```{r}
pie_crab %>% 
  slice_sample(n = 5)
```

Just to make things a little more interesting, I'm going to split up the dates into years, months, and days and save that as a new data frame, `crab_data`.

```{r}
crab_data <- pie_crab %>% 
  # extracting month from the date column using lubridate::month()
  # also making this a factor (instead of numeric) using as.factor()
  mutate(month = as.factor(month(date)))
```


# `ggplot` grammar

`ggplot` works in layers. The code to make a plot can vary, but always includes:  
1. the `ggplot()` call: this tells R that you want to use the function `ggplot()` in `ggplot` to plot things.  
2. `data` and `aesthetics` within that `ggplot()` call: tells `ggplot` to use a specific data frame any variables in that data frame that should be represented in the plot (for example, x- and y- axes, colors, shapes)  
3. a `geom_()`: short for "geometry", `geom_()` calls tell `ggplot` what kind of plot you want to make. Try `?geom_` in the console to see the different options.  

```{r}
# step 1: call ggplot
ggplot(
  
  # step 2: specify the data and the aesthetics
  # plotting latitude on the x-axis and crab size on the y-axis
  data = crab_data, aes(x = latitude, y = size)) +
  
  # step 3: specify a geom - in this case, we're creating a scatter plot
  geom_point()
```

Note that when you're adding on layers in `ggplot`, you'll use the `+` instead of the `%>%` operator. This is because `ggplot()` is the function call, but everything else you add on is a modifier of the `ggplot()` plotting function (instead of a new function doing something different).  

So we've just made this plot. But how can we make it _worse_?  

`ggplot()` takes aesthetics from the data frame, so I'm going to color the points by site and make the shapes represent month. I'm also going to make a jitter plot, which is a scatter plot with the points "jittered", or randomly shaken up so that it's easier to see the overlap (or be chaotic). I'm also going to facet the plot by month using `facet_wrap()`, which is a useful function when you're trying to see differences between variables in different panels (or you can use `facet_grid()`, which does essentially the same thing).
```{r}
ggplot(data = crab_data, aes(x = latitude, y = size)) +
  # putting the aesthetics in here: color points by site, shape points by month
  geom_jitter(aes(color = site, shape = month), 
  # anything that doesn't have to do with variables (like point size or transparency) goes outside the aesthetics
              size = 3, alpha = 0.6) +
  # facet by month
  facet_wrap(~ month)
```

Wow! This plot is getting worse to look at. But we can do better.

# Adjustments to `ggplot` defaults

## Manual selections
You can adjust the colors, shapes, axis limits, etc. using one of the `scale_` functions (try `?scale_` to see the many options).

## Themes
A lot of customization comes from `ggplot` themes. Theme elements deal with everything else in the plot that _doesn't_ have to do with the data (e.g. the plot background, grid, etc.). There are a lot of arguments that you can stack up in `theme()` (try `?theme()`) to see all of them, but they generally fall into a few categories: 1. axes, 2. legends, 3. panels, 4, plots, 5, strips. There are built-in themes in `ggplot` that can make your plot look pretty good right away, but playing around with theme elements yourself is a great way of getting the exact plot you want.  

Additionally, every argument takes a function modifying the _elements_: 1. lines, 2. rectangles (more broadly, shapes), 3. text. See the help pages for these functions (e.g. `?element_text()`) to figure out what you can alter about each element - there's a lot!  

## Annotations
Annotations are useful when you want to point something out on a plot. The function is `annotate()`, which adds an annotation layer onto your plot.

## Labels
Lastly, you can adjust any text on labels using the `labs()` function.  

Note: all the code in the theme below comes from Emily Riederer's gist, with some modifications!

```{r fig.height = 8, fig.width = 10}
ggplot(data = crab_data, aes(x = latitude, y = size)) +
  geom_jitter(aes(color = site, shape = month), size = 3, alpha = 0.6) +
  facet_wrap(~ month) +
  
  # scaling
  # changing point colors + assigning each to a site
  scale_color_manual(values = c("BC" = "#687f47", "CC" = "#268e86", "CT" = "#373f1f", "DB" = "#03010a", 
                                "GTM" = "#447c49", "JC" = "#2e2747", "NB" = "#7f696a", "NIB" = "#777a76", 
                                "PIE" = "#a87445", "RC" = "#aef23a", "SI" = "#e05357", "VCR" = "#087f0c", 
                                "ZI" = "#010201")) +
  # changing the point shapes
  scale_shape_manual(values = c("7" = 21, "8" = 18)) +
  # changing x and y axis limits and breaks
  scale_x_continuous(limits = c(29, 45)) +
  scale_y_continuous(n.breaks = 10) +

  # modifying the plot theme
  theme(
    
    # panel: anything having to do with the main area
    panel.background = element_rect(fill = '#B5C7F4', color = '#F761F9', linewidth = 5),
    panel.border = element_rect(fill = NA, color = "#87Bf18", linewidth = 2),
    panel.grid.major.x = element_line(color = "#FF21E1", linetype = 6),
    panel.grid.minor.x = element_line(color = "#2D65BF", linetype = 4),
    panel.grid.minor.y = element_blank(),
    
    # plot: anything having to do with the area around the panel
    plot.background = element_rect(fill = "#98FC6A"),
    plot.title = element_text(size = 30, hjust = 0.25, family = "Helvetica"),
    plot.subtitle = element_text(size = 20, hjust = 0.75, color = "#2A23A3", family = "Times New Roman"),
    plot.caption = element_text(size = 10, angle = 25, family = "Comic Sans MS"),
    plot.margin = unit(c(1, 4, 1, 3), "cm"), 
    
    # axes: anything having to do with the x- and y- axes
    axis.title.x = element_text(face = "bold.italic", color = "#9254D3"),
    axis.title.y = element_text(family = "Arial", face = "bold", size = 20, hjust = 0.25),
    axis.text = element_text(face = "italic", size = 15),
    axis.text.x.bottom = element_text(angle = 180), # note that axis.text options from above are inherited
    
    # strips: anything having to do with the facet titles
    strip.background = element_rect(fill = "#C9E886"),
    strip.text = element_text(color = "#854EED", family = "Garamond", face = "bold"),
    
    # legend: anything having to do with the legend
    legend.background = element_rect(fill = "#DC86E8"), 
    legend.key = element_rect(fill = "#C2F774"),
    legend.direction = "horizontal",
    legend.position = "top",
    legend.justification = "left",
    legend.title = element_text(family = "serif", color = "#B452F9"),
    legend.text = element_text(family = "mono", face = "italic", color = "#72FFFA")
    
  ) +
  
  # annotate: big gap
  annotate(geom = "text", x = 32, y = 20, label = "big gap", color = "#CC107D") +
  
  # labs: anything having to do with labels
  labs(title = "Crabs!",
       subtitle = "I love crabs",
       x = "Latitude (low to high)",
       y = "Size (big to small)",
       caption = "Do you love crabs?",
       col = "I love colors!") 
```

# Ugly plot contest

Now to create an ugly plot of your own!
