---
title: "Figure making in ggplot"
output: 
  html_document:
    toc: true
    toc_float: true
date: "13 February 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What we're doing today

In the spirit of Emily Riederer's [ugliest ggplot ever](https://gist.github.com/emilyriederer/2bf4f67d7e198f8359b61706c82e42ee), we'll play around with `ggplot` code in order to learn how it works. The goal: make the _ugliest_ plot possible.  

# Set up

We'll load in our packages below:
```{r message = FALSE, warning = FALSE}
# general use
library(tidyverse) # general tidying and visualization: ggplot is loaded by default with tidyverse
library(lterdatasampler) # data we're using comes from this package
library(lubridate) # working with dates
library(here) # folder organization

# extras
library(patchwork) # arranging plots
library(magick) # putting images into ggplots
```

Note: `lterdatasampler` has to be installed from the GitHub repo using the code below (copy, paste, and run in the console):
```{r eval = FALSE}
remotes::install_github("lter/lterdatasampler")
```

Today, we'll use the [Plum Island fiddler crab data](https://lter.github.io/lterdatasampler/articles/pie_crab_vignette.html) from `lterdatasampler` to visualize relationships between latitude and crab size. Read the linked vignette to learn about Bergmann's rule!  

The Plum Island LTER has a data set of crab sizes (column `size`) from Summer 2016 at 13 different marshes spanning 12 degrees latitude. A sample is below, but try `View(pie_crab)` in the console to see the whole data frame. 

```{r}
pie_crab %>% 
  slice_sample(n = 5)
```

Just to make things a little more interesting, I'm going to split up the dates into years, months, and days and save that as a new data frame, `crab_data`.

```{r}
crab_data <- pie_crab %>% 
  # extracting month from the date column using lubridate::month()
  # also making this a factor (instead of numeric) using as.factor()
  mutate(month = as.factor(month(date)))
```


# `ggplot` grammar

`ggplot` works in layers. The code to make a plot can vary, but always includes:  
1. the `ggplot()` call: this tells R that you want to use the function `ggplot()` in `ggplot` to plot things.  
2. `data` and `aesthetics` within that `ggplot()` call: tells `ggplot` to use a specific data frame any variables in that data frame that should be represented in the plot (for example, x- and y- axes, colors, shapes)  
3. a `geom_()`: short for "geometry", `geom_()` calls tell `ggplot` what kind of plot you want to make. Try `?geom_` in the console to see the different options.  

```{r}
# step 1: call ggplot

  
  # step 2: specify the data and the aesthetics
  # plotting latitude on the x-axis and crab size on the y-axis

  
  # step 3: specify a geom - in this case, we're creating a scatter plot

```

Note that when you're adding on layers in `ggplot`, you'll use the `+` instead of the `%>%` operator. This is because `ggplot()` is the function call, but everything else you add on is a modifier of the `ggplot()` plotting function (instead of a new function doing something different).  

So we've just made this plot. But how can we make it _worse_?  

`ggplot()` takes aesthetics from the data frame, so I'm going to color the points by site and make the shapes represent month. I'm also going to make a jitter plot, which is a scatter plot with the points "jittered", or randomly shaken up so that it's easier to see the overlap (or be chaotic). I'm also going to facet the plot by month using `facet_wrap()`, which is a useful function when you're trying to see differences between variables in different panels (or you can use `facet_grid()`, which does essentially the same thing).
```{r}

  # putting the aesthetics in here: color points by site, shape points by month

  # anything that doesn't have to do with variables (like point size or transparency) goes outside the aesthetics

  # facet by month

```

Wow! This plot is getting worse to look at. But we can do better.

# Adjustments to `ggplot` defaults

## Manual selections
You can adjust the colors, shapes, axis limits, etc. using one of the `scale_` functions (try `?scale_` to see the many options).

## Themes
A lot of customization comes from `ggplot` themes. Theme elements deal with everything else in the plot that _doesn't_ have to do with the data (e.g. the plot background, grid, etc.). There are a lot of arguments that you can stack up in `theme()` (try `?theme()`) to see all of them, but they generally fall into a few categories: 1. axes, 2. legends, 3. panels, 4, plots, 5, strips. There are built-in themes in `ggplot` that can make your plot look pretty good right away, but playing around with theme elements yourself is a great way of getting the exact plot you want.  

Additionally, every argument takes a function modifying the _elements_: 1. lines, 2. rectangles (more broadly, shapes), 3. text. See the help pages for these functions (e.g. `?element_text()`) to figure out what you can alter about each element - there's a lot!  

## Annotations
Annotations are useful when you want to point something out on a plot. The function is `annotate()`, which adds an annotation layer onto your plot.

## Labels
Lastly, you can adjust any text on labels using the `labs()` function.  

Note: all the code in the theme below comes from Emily Riederer's gist, with some modifications!

```{r fig.height = 8, fig.width = 10}
crab_plot <- ggplot(data = crab_data, aes(x = latitude, y = size)) +
  geom_jitter(aes(color = site, shape = month), size = 3, alpha = 0.6) +
  facet_wrap(~ month) +
  
  # scaling
  # 
  scale_color_manual(values = c("BC" = "#687f47", "CC" = "#268e86", "CT" = "#373f1f", 
                                "DB" = "#03010a", "GTM" = "#447c49", "JC" = "#2e2747", 
                                "NB" = "#7f696a", "NIB" = "#777a76", "PIE" = "#a87445", 
                                "RC" = "#aef23a", "SI" = "#e05357", "VCR" = "#087f0c", 
                                "ZI" = "#010201")) +
  # 
  scale_shape_manual(values = c("7" = 21, "8" = 18)) +
  # 
  scale_x_continuous(limits = c(29, 45)) +
  scale_y_continuous(n.breaks = 10) +
  
  # theme() call
  theme(
    # 
    panel.background = element_rect(fill = '#B5C7F4', color = '#F761F9', linewidth = 5),
    panel.border = element_rect(fill = NA, color = "#87Bf18", linewidth = 2),
    panel.grid.major.x = element_line(color = "#FF21E1", linetype = 6),
    panel.grid.minor.x = element_line(color = "#2D65BF", linetype = 4),
    panel.grid.minor.y = element_blank()
  ) +
  # another theme() call: you don't have to split things up like this 
  # just doing it for ease of demonstration
  theme(
    # 
    plot.background = element_rect(fill = "#98FC6A"),
    plot.title = element_text(size = 30, hjust = 0.25, family = "Helvetica"),
    plot.subtitle = element_text(size = 20, hjust = 0.75, color = "#2A23A3", family = "Times New Roman"),
    plot.caption = element_text(size = 10, angle = 25, family = "Comic Sans MS"),
    plot.margin = unit(c(1, 4, 1, 3), "cm")
  ) +
  theme(
    # 
    axis.title.x = element_text(face = "bold.italic", color = "#9254D3"),
    axis.title.y = element_text(family = "Arial", face = "bold", size = 20, hjust = 0.25),
    axis.text = element_text(face = "italic", size = 15),
    # note that axis.text options from above are inherited
    axis.text.x.bottom = element_text(angle = 180)
  ) +
  theme(
    # 
    strip.background = element_rect(fill = "#C9E886"),
    strip.text = element_text(color = "#854EED", family = "Garamond", face = "bold")
  ) +
  theme(
    # 
    legend.background = element_rect(fill = "#DC86E8"), 
    legend.key = element_rect(fill = "#C2F774"),
    legend.direction = "horizontal",
    legend.position = "top",
    legend.justification = "left",
    legend.title = element_text(family = "serif", color = "#B452F9"),
    legend.text = element_text(color = "#30F92C", family = "mono")
  ) +
  
  # 
  annotate(geom = "text", x = 32, y = 20, label = "big gap", color = "#CC107D") +
  
  # 
  labs(title = "Crabs!",
       subtitle = "I love crabs",
       x = "Latitude (low to high)",
       y = "Size (small to big)",
       caption = "Do you love crabs?",
       col = "I love colors!") 

crab_plot
```

The last thing we'll do is save the plot using `ggsave()`. The function needs you to tell it what the file name is in the first argument `filename`, but I'll show you a little trick I use to 1) save the figure in a folder and 2) "version" the figure based on date (i.e. the file name for the figure has the date it was made in it). Both are extraneous, but things I like to do to keep my figures organized.

```{r}
ggsave(
  # specify a file name, but here we'll specify a file path
             # choose the folder named "figures"
  filename = here::here("figures", 
                        # create a file name that includes the date using lubridate::today()
                        # you could also use Sys.Date() - works the same way!
                        paste("crab_plot_ugly_", today(), ".jpg", sep = "")),
  # name the plot object you're wanting to save
  plot = crab_plot,
  # specify the dimensions and units
  width = 10, height = 5, units = "in",
  # specify resolution
  dpi = 300
)
```

# Extras

## Adding images to plots

There are many ways of adding images to plots. The package I'll use is `magick`. This is a round-about way of getting an image in, but allows you to be flexible with your image types (gifs, etc.). I'll just save this as an object called `crab_image` to use later.

```{r}
# use magick::image_read() to read in the jpeg
crab_image <- image_read(here::here("code", "pwhittle-fiddler-crab.jpeg")) %>% 
  # turn this into a raster
  as.raster()
```

## Putting plots together

There are a couple packages you can use to put plots together like panels in a figure. The big ones are `patchwork` and `cowplot`. Try each of them out and see what you like, but we'll be using `patchwork` today.

Just to demonstrate how this works, I'm going to make up a histogram of crab sizes. The code is annotated if you're interested in the customized aesthetics, but they're not important for this demonstration.

```{r fig.height = 8, fig.width = 10}
# for a histogram, you don't need y in your aesthetic call
crab_hist <- ggplot(crab_data, aes(x = size)) +
  # telling ggplot that you want to plot a histogram
  geom_histogram(binwidth = 1, aes(fill = site)) +
  # some scaling
  scale_x_continuous(breaks = seq(from = 6.5, to = 24.5, by = 1)) +
  scale_y_continuous(limits = c(0, 45), expand = c(0, 0)) +
  scale_fill_manual(values = c("BC" = "#7D9084", "CC" = "#8B9FD9", "CT" = "#DDD5D5", 
                               "DB" = "#CADDA5", "GTM" = "#DF697F", "JC" = "#E09E5F", 
                               "NB" = "#7F56DB", "NIB" = "#D8DF5D", "PIE" = "#DC53D3", 
                               "RC" = "#84D9DF", "SI" = "#D798D1", "VCR" = "#77E2A8", 
                               "ZI" = "#76E762")) +
  # some theme-ing
  theme(
    panel.background = element_rect(fill = '#B6E45F', color = '#DABA84', linewidth = 5),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#E16B86", linetype = 2),
    panel.grid.minor.y = element_line(color = "#B49CD2", linetype = 3),
    legend.background = element_rect(fill = "#A0C8E8"),
    plot.background = element_rect(fill = "#8ADDA8"),
    plot.title = element_text(size = 20),
    plot.margin = unit(c(1, 4, 1, 3), "cm")
  ) +
  # some labeling
  labs(x = "Crab sizes (small to big)",
       y = "Number of crabs (few to lots)",
       title = "This is a crab histogram",
       fill = "Places") +
  # adding the crab image
  annotation_raster(crab_image, xmin = 8, xmax = 10, ymin = 30, ymax = 35) +
  # annotating the plot with the photographer credit
  annotate(geom = "text", x = 9, y = 29, label = "PWhittle, iNaturalist", size = 3)
  
crab_hist
```

`patchwork` thinks similarly to `ggplot` in that it takes things in layers with `+` signs, but also takes `&`. To be honest, I can't remember the rules most of the time and I have to look them up.
```{r fig.height = 8, fig.width = 18}
                       # this automatically recognizes that you're putting plots together (no function call!)
crab_plots_together <- crab_plot + crab_hist + 
  
  # plot_layout: anything having to do with the way plots are arranged
  # widths takes a ratio: the panel ratio will be 2:1 left:right
  plot_layout(widths = c(2, 1)) + 
  
  # plot_annotation: anything having to do with annotations
  plot_annotation(tag_levels = "A", tag_suffix = ")") & # annotating them with panel letters
  
  # patchwork has its own theme options!
  theme(plot.tag = element_text(size = 40)) # changing the size of the annotations

# display our beautiful plot  
crab_plots_together

# saving the plot
ggsave(
  filename = here::here("figures", 
                        paste("crab_plot_and_hist_", today(), ".jpg", sep = "")),
  plot = crab_plots_together,
  width = 18, height = 8, units = "in",
  dpi = 150
)
```




# Ugly plot contest

Now to create an ugly plot of your own!
